Enzo: An Adaptive Mesh Refinement Code for Astrophysics

### 2013-06-24 introduction.tex
Changes from Cameron Hummels comments (also some other typos I found while working through his comments).

The \enzo\ code has been extensively used over the last two decades in
a wide variety of problems, resulting in the publication of more than
100 peer-reviewed papers based on results obtained with the code. The
variety of astrophysical systems that \enzo\ has been used for include
galaxies \citep{2003ApJ...587...13T, 2012MNRAS.425..641L,
  2012ApJ...749..140H, 2013MNRAS.432.1989S}, galaxy
clusters \citep{Loken02, Xu11, Skillman13}, the interstellar medium
\citep{Slyz05}, the intergalactic medium \citep{Fang01,
2011ApJ...731....6S}, cooling flows \citep{Li12, 2013ApJ...763...38S},
turbulence \citep{Kritsuk04, 2007ApJ...665..416K, 2009JPhCS.180a2020K, Collins11}, the
formation of the first stars \citep{ABN02, 2007ApJ...654...66O,
2009Sci...325..601T, Xu08}, and the formation of stars in our own Galaxy
\citep{Collins11, 2011ApJ...727L..20K, Collins12a}.

>>>>

The \enzo\ code has been extensively used over the last two decades in
a wide variety of problems, resulting in the publication of more than
100 peer-reviewed papers based on results obtained with the code. The
variety of astrophysical systems that \enzo\ has been used for include
galaxies \citep{2003ApJ...587...13T, 2012MNRAS.425..641L,
  2012ApJ...749..140H, 2013MNRAS.432.1989S}, galaxy
clusters \citep{Loken02, Xu11, Skillman13}, the interstellar medium
\citep{Slyz05}, the intergalactic medium \citep{Fang01, 
2011ApJ...731....6S}, the circumgalactic
medium~\citep{2013MNRAS.430.1548H}, cooling flows \citep{Li12, 2013ApJ...763...38S},
turbulence \citep{Kritsuk04, 2007ApJ...665..416K, 2009JPhCS.180a2020K, Collins11}, the
formation of the first stars \citep{ABN02, 2007ApJ...654...66O,
2009Sci...325..601T, Xu08}, and the formation of stars in our own Galaxy
\citep{Collins11, 2011ApJ...727L..20K, Collins12a}.

### 2013-06-24 numerical-amr.tex
Changes from Cameron Hummels comments (also some other typos I found while working through his comments).

The back-bone of the SAMR idea is the patch, or grid, which we take to be rectilinear to simplify bookkeeping (in practice, the lower-level routines such as the hydro solvers all assume Cartesian coordinates).  The single root grid covers the entire computational domain and plays the role of the root node in the grid hierarchy.  Subgrids are placed such that they cover (hopefully small) sub-volumes of the root grid at higher resolution.  This structure forms a tree that may be extended to arbitrary depths by adding additional grids to refine regions in the subgrids.  We use the following notation to describe grid relationships.  A grid's \textit{parent} is the coarse grid that completely contains it.  The \textit{root grid} is the most coarsely-resolved grid, and has no parent.  A \textit{child grid} is grid completely contained by its parent grid (note that a grid may have many children but only one parent), while sibling grids are those that have the same resolution, but not necessarily the same parent.  We also refer to levels of the hierarchy, where the root grid is labeled level zero and all grids with the same resolution are on the
same level.

@@

In the second case, the value must be either (1) interpolated from a parent, or (2) copied from a sibling.  In practice, we solve this problem by first interpolating all boundaries from the grid's parent and then performing the sibling copy wherever possible, since these values will always be as good as, or superior to, interpolation.

@@

In three dimensions this process breaks down since the number of
constraints (center + four diagonals = five) exceeds the number of
parameters (four).  One way to solve this problem would be to perform
a least-squares fit under the given constraints, however, this is
computationally too expensive.  Therefore, in Appendix~\ref{app:interpolation} we describe
a heuristic approach that satisfies four constaints analytically and
then adjusts the interpolation parameters in such a way as to satisfy the fifth.
While the result is not necessarily optimal, it does appear to perform
adequately in our tests.

>>>>

The back-bone of the SAMR idea is the patch, or grid, which we take to be rectilinear to simplify bookkeeping (in practice, the lower-level routines such as the hydro solvers all assume Cartesian coordinates).  The single root grid covers the entire computational domain and plays the role of the root node in the grid hierarchy.  Subgrids are placed such that they cover (hopefully small) sub-volumes of the root grid at higher resolution.  This structure forms a tree that may be extended to arbitrary depths by adding additional grids to refine regions in the subgrids.  We use the following notation to describe grid relationships.  A grid's \textit{parent} is the coarse grid that completely contains it.  The \textit{root grid} is the most coarsely-resolved grid, and it has no parent.  A \textit{child grid} is a grid completely contained by its parent grid (note that a grid may have many children but only one parent), while sibling grids are those that have the same resolution, but not necessarily the same parent.  We also refer to levels of the hierarchy, where the root grid is labeled level zero and all grids with the same resolution are on the
same level.

@@

In the second case, the value must be either (1) interpolated from a parent or (2) copied from a sibling.  In practice, we solve this problem by first interpolating all boundaries from the grid's parent and then performing the sibling copy wherever possible, since these values will always be as good as, or superior to, interpolation.

@@

In three dimensions this process breaks down since the number of
constraints (center + four diagonals = five) exceeds the number of
parameters (four).  One way to solve this problem would be to perform
a least-squares fit under the given constraints, however, this is
too computationally expensive.  Therefore, in Appendix~\ref{app:interpolation} we describe
a heuristic approach that satisfies four constaints analytically and
then adjusts the interpolation parameters in such a way as to satisfy the fifth.
While the result is not necessarily optimal, it does appear to perform
adequately in our tests.

### 2013-06-24 numerical-chemistry.tex
Changes from Cameron Hummels comments (also some other typos I found while working through his comments).

The primordial non-equilibrium chemistry routines used in \enzo\ were first
described by \citet{abel97} and \citet{anninos97}, but have
since been extended with updated reaction rates and the inclusion of
deuterium species \citep{2009PhDT.........5T}.  These routines follow the
non-equilibrium chemistry of a gas of primordial composition with 12 total
species: H, \Hp, He, \Hep, \Hepp, \Hm, \HHp, \HH, e$^-$, D, \Dp, and HD.  \enzo\
also computes the radiative heating and cooling of the gas from atomic and
molecular line excitation, recombination, collisional excitation, free-free
transitions, Compton scattering of the cosmic microwave background, as well as
several models for a metagalactic UV background that heat the gas via
photoionization and photodissociation; see Section~\ref{sec.num.cooling} for
more details.  The chemical and thermal states of the gas can be updated either
at the same hydrodynamical timestep (i.e., de-coupled and operator split) or
through the same subcycling system (i.e., a coupled chemical and thermal
system).  The default behavior of \enzo\ is to couple these two systems at
subcycles of the hydrodynamic timestep; this results in updates to both the
chemical and thermal states of the gas (which also inform the temperature, the
reaction rate coefficients and the cooling functions of the gas) on timescales
that are faster than those of the gas dynamics.  The gamma used by \enzo\ to
compute the temperature of the gas from the energy and density characteristics
utilizes a variable gamma that includes effects of the rotational state of
molecular hydrogen, enabling it to vary from $5/3$ (fully-atomic) to $7/5$
(fully molecular).  This further coupling of the chemical and thermal states of
the gas underscores the need for coupled chemistry and radiative cooling
solutions.

>>>>

The primordial non-equilibrium chemistry routines used in \enzo\ were first
described by \citet{abel97} and \citet{anninos97}, but have
since been extended with updated reaction rates and the inclusion of
deuterium species \citep{2009PhDT.........5T}.  These routines follow the
non-equilibrium chemistry of a gas of primordial composition with 12 total
species: H, \Hp, He, \Hep, \Hepp, \Hm, \HHp, \HH, e$^-$, D, \Dp, and HD.  \enzo\
also computes the radiative heating and cooling of the gas from atomic and
molecular line excitation, recombination, collisional excitation, free-free
transitions, Compton scattering of the cosmic microwave background, as well as
several models for a metagalactic UV background that heat the gas via
photoionization and photodissociation; see Section~\ref{sec.num.cooling} for
more details.  The chemical and thermal states of the gas can be updated either
at the same hydrodynamical timestep (i.e., de-coupled and operator-split) or
through the same subcycling system (i.e., a coupled chemical and thermal
system).  The default behavior of \enzo\ is to couple these two systems at
subcycles of the hydrodynamic timestep; this results in updates to both the
chemical and thermal states of the gas (which also inform the temperature, the
reaction rate coefficients and the cooling functions of the gas) on timescales
that are faster than those of the gas dynamics.  The gamma used by \enzo\ to
compute the temperature of the gas from the energy and density characteristics
utilizes a variable gamma that includes effects of the rotational state of
molecular hydrogen, enabling it to vary from $5/3$ (fully-atomic) to $7/5$
(fully molecular).  This further coupling of the chemical and thermal states of
the gas underscores the need for coupled chemistry and radiative cooling
solutions.

### 2013-06-24 numerical-gravity.tex
Changes from Cameron Hummels comments (also some other typos I found while working through his comments).

% Changed:  ``Particles from sub-grids are also added in this way'' to ``Particles on sub-grids within the grid's volume are also added to its gravitating field using the same method.'' I expanded because I missed the point and had to clarify with Brian when I read the text. But if it's too long, please just cut it!

@@

First, the massive (dark matter and star) particles are distributed onto the grids using the second-order cloud-in-cell (CIC) interpolation technique \citep{Hockney88} to form a spatially-discretized density field $\rho_{\rm DM}$.  During the CIC interpolation, particle positions are (temporarily) advanced by $0.5 \Delta t v^n$ so that we generate an estimate of the time-centered density field.  Particles on sub-grids within the grid's volume are also added to its gravitating field using the same method. In addition, since the gravitating field for a grid is defined beyond the grid edges (see below), massive particles from sibling grids and sub grids that lie within the entire gravitating field are used.  This step can involve communication.

@@

In order to calculate more accurate potentials on the subgrids, \enzo\ uses a similar but slightly different technique from the root grid.  The generation of the total gravitating mass field is essentially identical, using CIC interpolation for both the particles and baryons, including sub-grids as before.  To compute the potential on subgrids, however, we use the standard seven-point (in three dimensions) second-order finite difference approximation to Poisson's equation.  Boundary conditions are then interpolated from the potential values on the parent grid.  We use either tri-linear interpolation or a natural second-order spline for this: both methods give similar results but the default is the tri-linear interpolation, which empirically provides a resonable compromise between speed and accuracy. The potential equation on each subgrid is then solved with the given Dirichlet boundary conditions with a multigrid relaxation technique.  This is applied to each subgrid separately.

>>>>

% Changed:  ``Particles from subgrids are also added in this way'' to ``Particles on subgrids within the grid's volume are also added to its gravitating field using the same method.'' I expanded because I missed the point and had to clarify with Brian when I read the text. But if it's too long, please just cut it!

@@

First, the massive (dark matter and star) particles are distributed onto the grids using the second-order cloud-in-cell (CIC) interpolation technique \citep{Hockney88} to form a spatially-discretized density field $\rho_{\rm DM}$.  During the CIC interpolation, particle positions are (temporarily) advanced by $0.5 \Delta t v^n$ so that we generate an estimate of the time-centered density field.  Particles on subgrids within the grid's volume are also added to its gravitating field using the same method. In addition, since the gravitating field for a grid is defined beyond the grid edges (see below), massive particles from sibling grids and sub grids that lie within the entire gravitating field are used.  This step can involve communication.

@@

In order to calculate more accurate potentials on the subgrids, \enzo\ uses a similar but slightly different technique from the root grid.  The generation of the total gravitating mass field is essentially identical, using CIC interpolation for both the particles and baryons, including subgrids as before.  To compute the potential on subgrids, however, we use the standard seven-point (in three dimensions) second-order finite difference approximation to Poisson's equation.  Boundary conditions are then interpolated from the potential values on the parent grid.  We use either tri-linear interpolation or a natural second-order spline for this: both methods give similar results, but the default is the tri-linear interpolation, which empirically provides a resonable compromise between speed and accuracy. The potential equation on each subgrid is then solved with the given Dirichlet boundary conditions with a multigrid relaxation technique.  This is applied to each subgrid separately.

### 2013-06-24 numerical-mhd-ct.tex
Changes from Cameron Hummels comments (also some other typos I found while working through his comments).

The increased complexity of the constrained transport scheme comes in
the form of area-averaged face and length-averaged edge-centered
variables, while the rest of \enzo\ employs predominantly
volume-averaged cell centered variables.  The magnetic field is
represented by both a face centered field, B$_f$, and a cell centered
field, B$_c$.  The electric field is edge centered.  The magnetic
field is updated in four steps: first, the Riemann problem is solved
in the traditional manner, using the cell centered field; second, an
edge-centered electric field is computed using the fluxes from the
Riemann solver; third, the curl of that electric field is used to
update the face centered field; finally, the cell centered magnetic
field is updated with an average of the face centered field.

>>>>

The increased complexity of the constrained transport scheme comes in
the form of area-averaged face and length-averaged edge-centered
variables, while the rest of \enzo\ employs predominantly
volume-averaged cell-centered variables.  The magnetic field is
represented by both a face-centered field, B$_f$, and a cell-centered
field, B$_c$.  The electric field is edge-centered.  The magnetic
field is updated in four steps: first, the Riemann problem is solved
in the traditional manner, using the cell-centered field; second, an
edge-centered electric field is computed using the fluxes from the
Riemann solver; third, the curl of that electric field is used to
update the face-centered field; finally, the cell-centered magnetic
field is updated with an average of the face-centered field.

### 2013-06-24 numerical-mhd-dedner.tex
Changes from Cameron Hummels comments (also some other typos I found while working through his comments).

To more clearly describe the Dedner divergence cleaning modifications, we write the equations of compressible inviscid hydrodynamics can be written in the form of conservation laws as,
\begin{equation}
 \frac{\partial{U}}{\partial{t}} +
 \frac{\partial{F^x}}{\partial{x}} + \frac{\partial{F^y}}{\partial{y}} + \frac{\partial{F^z}}{\partial{z}}= 0, \label{hydro}
\end{equation}
The conserved variable $U$ is given by
\begin{equation}
 U = (\rho, \rho v_x, \rho v_y, \rho v_z, \rho E)^{T},
\end{equation} 
where $\rho$ is density, $v_i$ are the three components of velocity
for $i={x,y,z}$, $E=v^2/2 + e$ denotes the specific total energy and $e$ the
specific internal energy.

>>>>

To more clearly describe the Dedner divergence cleaning modifications, we write the equations of compressible inviscid hydrodynamics in the form of conservation laws as,
\begin{equation}
 \frac{\partial{U}}{\partial{t}} +
 \frac{\partial{F^x}}{\partial{x}} + \frac{\partial{F^y}}{\partial{y}} + \frac{\partial{F^z}}{\partial{z}}= 0, \label{hydro}
\end{equation}
The conserved variable $U$ is given by
\begin{equation}
 U = (\rho, \rho v_x, \rho v_y, \rho v_z, \rho E)^{T},
\end{equation} 
where $\rho$ is density, $v_i$ are the three components of velocity
for $i={x,y,z}$, $E=v^2/2 + e$ denotes the specific total energy and $e$ the
specific internal energy.

### 2013-06-24 numerical-overview.tex
Changes from Cameron Hummels comments (also some other typos I found while working through his comments).

PPM is an explicit, higher order-accurate version of
Godunov's method for ideal gas dynamics with third-order-accurate piecewise parabolic
monotonic interpolation and a nonlinear Riemann solver for shock
capturing.  It advances the hydrodynamic equations in the following
steps:
\begin{enumerate}
 \item Construct monotonic parabolic interpolation of cell average
 data, for each fluid quantity.
 \item Compute interface states by averaging the parabola for over the domain of dependence for
 each interface
 \item Use interface data to solve the Riemann problem.
 \item Difference the interface fluxes to update the cell average quantities.
\end{enumerate}
It does an excellent job of capturing strong shocks in at
most two cells.  Multidimensional schemes are built up by directional
splitting and produce a method that is formally second-order-accurate
in space and time which explicitly conserves mass, linear momentum, and energy.

@@

As a check on PPM, \enzo\ also includes an implementation of the
finite difference hydrodynamic algorithm employed in the compressible 
magnetohydrodynamics code `\zeus'
\citep{Stone92a, Stone92b}.  Fluid transport is solved on a Cartesian
grid using the upwind, monotonic advection scheme of \citet{1977JCoPh..23..276V} 
within a multistep 
(operator split) solution procedure which is fully explicit in time.  
This method is formally second-order-accurate in space but 
first-order-accurate in time.  It is important to note that 
this method conserves internal energy
rather than total energy, so the ``dual energy
formulation'' discussed in Section~\ref{sec.ov.hydro.ppm} is unnecessary.

Operator split methods break
the solution of the hydrodynamic equations into parts, with each part
representing a single term in the equations.  Each part is evaluated
successively using the results preceding it.  The individual parts of
the solution are grouped into two steps, called the \emph{source} and
\emph{transport} steps.

>>>>

PPM is an explicit, higher-order accurate version of
Godunov's method for ideal gas dynamics with third-order accurate piecewise parabolic
monotonic interpolation and a nonlinear Riemann solver for shock
capturing.  It advances the hydrodynamic equations in the following
steps:
\begin{enumerate}
 \item Construct monotonic parabolic interpolation of cell average
 data, for each fluid quantity.
 \item Compute interface states by averaging the parabola for over the domain of dependence for
 each interface
 \item Use interface data to solve the Riemann problem.
 \item Difference the interface fluxes to update the cell average quantities.
\end{enumerate}
It does an excellent job of capturing strong shocks in at
most two cells.  Multidimensional schemes are built up by directional
splitting and produce a method that is formally second-order accurate
in space and time which explicitly conserves mass, linear momentum, and energy.

@@

As a check on PPM, \enzo\ also includes an implementation of the
finite difference hydrodynamic algorithm employed in the compressible 
magnetohydrodynamics code `\zeus'
\citep{Stone92a, Stone92b}.  Fluid transport is solved on a Cartesian
grid using the upwind, monotonic advection scheme of \citet{1977JCoPh..23..276V} 
within a multistep 
(operator-split) solution procedure which is fully explicit in time.  
This method is formally second-order accurate in space but 
first-order accurate in time.  It is important to note that 
this method conserves internal energy
rather than total energy, so the ``dual energy
formulation'' discussed in Section~\ref{sec.ov.hydro.ppm} is unnecessary.

Operator-split methods break
the solution of the hydrodynamic equations into parts, with each part
representing a single term in the equations.  Each part is evaluated
successively using the results preceding it.  The individual parts of
the solution are grouped into two steps, called the \emph{source} and
\emph{transport} steps.

### 2013-06-24 numerical-particle.tex
Changes from Cameron Hummels comments (also some other typos I found while working through his comments).

Particles are stored in the most highly refined grid patch at the
point in space where they exist, and particles that move out of a
subgrid patch are sent to the grid patch covering the adjacent volume
with the finest spatial resolution, which may be of the same spatial
resolution, coarser, or finer than the grid patch that the particles
are moved from.  This takes place in a communication process at the
end of each timestep on a level.

To avoid unphysical point-mass effects, \enzo\ provides a parameter that
governs the maximum level at which particles will be regarded at point masses.
At higher levels, contributions from particles to the gravitating mass field
will be smoothed over a spherical region centered at each particle's center.

>>>>

Particles are stored in the most highly refined grid patch at the
point in space where they exist, and particles that move out of a
subgrid patch are sent to the grid patch covering the adjacent volume
with the finest spatial resolution, which may be of the same spatial
resolution, coarser, or finer than the grid patch from which the particles
moved.  This takes place in a communication process at the
end of each timestep on a level.

To avoid unphysical point-mass effects, \enzo\ provides a parameter that
governs the maximum level at which particles will be regarded as point masses.
At higher levels, contributions from particles to the gravitating mass field
will be smoothed over a spherical region centered at each particle's center.

### 2013-06-24 numerical-ppm.tex
Changes from Cameron Hummels comments (also some other typos I found while working through his comments).

One (purely) hydrodynamic method used in \enzo\ is closely based on the
piecewise parabolic method (PPM) of
\citet{1984JCoPh..54..174C}, which has been
modified for the study of cosmological fluid flows.  
\citet{1984JCoPh..54..174C} describe two variants on this method:
Lagrangian Remap and Direct Eulerian; here we use the Direct Eulerian
version which is better suited for AMR simulations.  The Lagrangian
Remap version has previously been adapted for cosmological use
as described in \citet{1995CoPhC..89..149B} and the implementation
we use here is closely based on that work.

The first term on the right hand side of equations~(\ref{eq:momentum}) and
(\ref{eq:total_energy}) come from our choice of comoving coordinates
(a similar term also appears in equation~(\ref{eq:dm_velocity}), the velocity 
relation for the dark matter particles).  A similar term does not
appear in the mass conservation equation (\ref{eq:mass}) because
of the comoving density definition.   
We note that these expansion terms could be
eliminated entirely by the proper choice of variables (including time),
although we have not done so here, as they do not constitute a major
source of error.  We solve these terms
using the same technique -- they are split from the rest of the terms and
solved using an (implicit) time-centered method, which is straightforward
as there are no spatial gradients.  Note that we use this method for
all four of the hydro solvers.

@@

We now restrict ourselves to the solution of equations
(\ref{eq:mass1d})--(\ref{eq:total_energy1d}), in one
dimension.  In the Direct Eulerian version of PPM, this is
accomplished by a three-step process.  First, we compute effective
left and right states at each grid boundary by constructing a
piecewise parabolic description of the primative variables ($\rho$, $u$
and $E$) and then averaging over the regions corresponding to the
distance each characteristic wave can travel ($u$, $u-c_s$ and
$u+c_s$, where $c_s$ is the sound speed).
Second, a Riemann problem is solved using these effective left and
right states, and finally the fluxes are computed based on the
solution to this Riemann problem and the conserved quantities are
updated.  This is described in detail in
\citet{1984JCoPh..54..174C}, but we will briefly outline the procedure
here both for completeness and to put the changes we will make in
context.

@@

We first construct monotonic piecewise 
parabolic (third-order) interpolations in one dimension
for each of $p$, $\rho$, and $v$.  The pressure is determined from
equation~(\ref{eq:eq_of_state}), the equation of state.
The interpolation formula for some quantity $q$ is given by:
%
\begin{eqnarray}
q_j(x) & = &  q_{L,j} + \tilde{x}(\Delta q_j + q_{6,j}(1-\tilde{x})), \\
\tilde{x}      & \equiv & {x - x_{j-1/2} \over \Delta x_j}, \qquad
             x_{j-1/2} \leq x \leq x_{j+1/2}. \nonumber
\end{eqnarray}
%
This is equation~(1.4) of CW84 (in the spatial rather than mass coordinate as is appropriate for the direct Eulerian approach). 
The quantities $q_{L,j}$, $\Delta q_j$,
and $q_{6,j}$ can be viewed simply as interpolation constants; however,
they also have more intuitive meanings.  For example, $q_{L,j}$ is the
value of $q$ at the left edge of zone j, while $\Delta q_j$ and $q_{6,j}$ are analogous to the slope and first-order correction to the slope of $q$ (see \citet{1984JCoPh..54..174C} for a complete discussion):
\begin{equation}
\Delta q_j \equiv q_{R,j} - q_{L,j} \qquad 
q_{6,j}    \equiv 6\left[q_j - 1/2\left(q_{L,j} + q_{R,j}\right)\right].
\end{equation}

We have reduced the problem to finding $q_{L,j}$ and $q_{R,j}$.  While this
is simple in principle, it is complicated somewhat by the requirement that
these values be of sufficient accuracy and that the resulting distribution
be monotonic.  That is, no new maxima or minima are introduced.
The resulting formulae are straightforward but complicated and are not
reproduced here, but see Equations 1.7 to 1.10 of \citet{1984JCoPh..54..174C}.
We also optionally allow steepening as described in that reference.

Once we have the reconstruction, the primary quantities ($p$, $\rho$, $v$ and $E$) are averaged over the domains corresponding to the three characteristics $C = u-c$, $u$ or $u+c$ (where $c$ is the sound speed in a cell):
\begin{equation}
      q^{C+}_{j+1/2} =
          {1 \over \Delta t C^n_j} \int^{x_{j+1/2}}_{x_{j+1/2}-\Delta t C^n_j} q_j(x) dx
\end{equation}
where $q^{C+}$ refers to the left characteristic $C$ (and a similar integral is done for the right characteristic $q^{C-}$).  The linearized gas dynamics equations are then used to compute second-order accurate left and right states that take into account the multiple wave families.  This process is described in \citet{1984JCoPh..54..174C} and we use their equations (3.6) and (3.7).

@@

We include an optional diffusive flux (and flattening for the parabolic curves) that can improve the solution in some cases.  Our implementation follows that in the appendix of \citet{1984JCoPh..54..174C}.

In addition, as discussed earlier, the three dimensional scheme is achieved by operator splitting and repeating the above procedure in the other two orthogonal directions.  The transverse velocities and any additional passive quantities are naturally and easily added to this system (see Equation 3.6 of \citet{1984JCoPh..54..174C}).

We note that the acceleration required in Equation~(\ref{eq:momentum_diff}) is actually the
acceleration felt by the entire zone and not just at the zone center.
Therefore, it is possible to find the mass weighted average acceleration over the zone
by expanding the density and acceleration distributions and
retaining all terms up to second-order in $\Delta x$:
%
\begin{equation}
g_j^{n+1/2} = 
       \frac{1}{2 a^{n+1/2} \delta x_j} \left[ 
             \phi_{j+1}^{n+1/2} 
           - \phi_{j-1}^{n+1/2} 
           + \frac{1}{12} \left(    \phi^{n+1/2}_{j+1} 
                                 - 2\phi^{n+1/2}_j 
                                 + \phi^{n+1/2}_{j-1} \right) 
                                   {\delta d_j \over d_j}
       \right].
       \label{eq:accel}
\end{equation}
Generally we find that the potential is so slowly varying that this change is unnecessary.

@@

While there exists an exact solution to this problem, finding it is expensive.  There are four
possible combination of left- and right- traveling shocks and rarefactions,
only one of which is fully consistent with the initial conditions.  Once the
correct physical state is determined, the pressure in the central region can
only be found by finding the root to an algebraic equation, which is necessarily
an iterative process.  Thus a series of \emph{approximate} Riemann solvers are
typically used.  There are four approximate Riemann solvers in \enzo: two-shock
\citep{toro-1997},
Harten-Lax-van Leer \citep[HLL,][]{toro-1997}, HLL with a contact discontinuity
 \citep[HLLC,][]{toro-1997}, and HLL with
multiple discontinuities \citep[HLLD,][]{Miyoshi05}.  Two-shock is used only
with the PPM method.  HLL and HLLC are used with PPM, MUSCL (both with and
without MHD) and MHD-CT.  HLLD is exclusively an MHD solver, and works with both
MUSCL and MHD-CT methods.

The only approximation that two-shock makes is that both left- and right- moving
waves are shocks.  This solution still requires an iterative method for finding
the pressure in between the two waves.  The HLL method alleviates this iteration
by assuming that there is no central contact discontinuity, and the signal speed
in the central region is approximated by an average over the left- and right-
moving waves.  This method is significantly faster than the two-shock method,
but also quite a bit more dissipative.  The HLLC is a three-wave method that improves upon the HLL
method by also including the third wave, the contact discontinuity.

>>>>

One (purely) hydrodynamic method used in \enzo\ is closely based on the
piecewise parabolic method (PPM) of
\citet{1984JCoPh..54..174C} -- henceforth referred to as CW84 -- which has been
modified for the study of cosmological fluid flows.  
CW84 describe two variants on this method:
Lagrangian Remap and Direct Eulerian; here we use the Direct Eulerian
version which is better suited for AMR simulations.  The Lagrangian
Remap version has previously been adapted for cosmological use
as described in \citet{1995CoPhC..89..149B}, and the implementation
we use here is closely based on that work.

The first term on the right hand side of equations~(\ref{eq:momentum}) and
(\ref{eq:total_energy}) comes from our choice of comoving coordinates
(a similar term also appears in equation~(\ref{eq:dm_velocity}), the velocity 
relation for the dark matter particles).  A similar term does not
appear in the mass conservation equation (\ref{eq:mass}) because
of the comoving density definition.   
We note that these expansion terms could be
eliminated entirely by the proper choice of variables (including time),
although we have not done so here, as they do not constitute a major
source of error.  We solve these terms
using the same technique -- they are split from the rest of the terms and
solved using an (implicit) time-centered method, which is straightforward
as there are no spatial gradients.  Note that we use this method for
all four of the hydro solvers.

@@

We now restrict ourselves to the solution of equations
(\ref{eq:mass1d})--(\ref{eq:total_energy1d}), in one
dimension.  In the Direct Eulerian version of PPM, this is
accomplished by a three-step process.  First, we compute effective
left and right states at each grid boundary by constructing a
piecewise parabolic description of the primative variables ($\rho$, $u$
and $E$) and then averaging over the regions corresponding to the
distance each characteristic wave can travel ($u$, $u-c_s$ and
$u+c_s$, where $c_s$ is the sound speed).
Second, a Riemann problem is solved using these effective left and
right states, and finally the fluxes are computed based on the
solution to this Riemann problem and the conserved quantities are
updated.  This is described in detail in
CW84, but we will briefly outline the procedure
here both for completeness and to put the changes we will make in
context.

@@

We first construct monotonic piecewise 
parabolic (third-order) interpolations in one dimension
for each of $p$, $\rho$, and $v$.  The pressure is determined from
equation~(\ref{eq:eq_of_state}), the equation of state.
The interpolation formula for some quantity $q$ is given by:
%
\begin{eqnarray}
q_j(x) & = &  q_{L,j} + \tilde{x}(\Delta q_j + q_{6,j}(1-\tilde{x})), \\
\tilde{x}      & \equiv & {x - x_{j-1/2} \over \Delta x_j}, \qquad
             x_{j-1/2} \leq x \leq x_{j+1/2}. \nonumber
\end{eqnarray}
%
This is equation~(1.4) of CW84 (in the spatial rather than mass coordinate as is appropriate for the direct Eulerian approach). 
The quantities $q_{L,j}$, $\Delta q_j$,
and $q_{6,j}$ can be viewed simply as interpolation constants; however,
they also have more intuitive meanings.  For example, $q_{L,j}$ is the
value of $q$ at the left edge of zone j, while $\Delta q_j$ and $q_{6,j}$ are analogous to the slope and first-order correction to the slope of $q$ (see CW84 for a complete discussion):
\begin{equation}
\Delta q_j \equiv q_{R,j} - q_{L,j} \qquad 
q_{6,j}    \equiv 6\left[q_j - 1/2\left(q_{L,j} + q_{R,j}\right)\right].
\end{equation}

We have reduced the problem to finding $q_{L,j}$ and $q_{R,j}$.  While this
is simple in principle, it is complicated somewhat by the requirement that
these values be of sufficient accuracy and that the resulting distribution
be monotonic.  That is, no new maxima or minima are introduced.
The resulting formulae are straightforward but complicated and are not
reproduced here, but see Equations 1.7 to 1.10 of CW84.
We also optionally allow steepening as described in that reference.

Once we have the reconstruction, the primary quantities ($p$, $\rho$, $v$ and $E$) are averaged over the domains corresponding to the three characteristics $C = u-c$, $u$ or $u+c$ (where $c$ is the sound speed in a cell):
\begin{equation}
      q^{C+}_{j+1/2} =
          {1 \over \Delta t C^n_j} \int^{x_{j+1/2}}_{x_{j+1/2}-\Delta t C^n_j} q_j(x) dx
\end{equation}
where $q^{C+}$ refers to the left characteristic $C$ (and a similar integral is done for the right characteristic $q^{C-}$).  The linearized gas dynamics equations are then used to compute second-order accurate left and right states that take into account the multiple wave families.  This process is described in CW84 and we use their equations (3.6) and (3.7).

@@

We include an optional diffusive flux (and flattening for the parabolic curves) that can improve the solution in some cases.  Our implementation follows that in the appendix of CW84.

In addition, as discussed earlier, the three dimensional scheme is achieved by operator-zsplitting and repeating the above procedure in the other two orthogonal directions.  The transverse velocities and any additional passive quantities are naturally and easily added to this system (see Equation 3.6 of CW84).

We note that the acceleration required in Equation~(\ref{eq:momentum_diff}) is actually the
acceleration felt by the entire zone and not just at the zone center.
Therefore, it is possible to find the mass-weighted average acceleration over the zone
by expanding the density and acceleration distributions and
retaining all terms up to second-order in $\Delta x$:
%
\begin{equation}
g_j^{n+1/2} = 
       \frac{1}{2 a^{n+1/2} \delta x_j} \left[ 
             \phi_{j+1}^{n+1/2} 
           - \phi_{j-1}^{n+1/2} 
           + \frac{1}{12} \left(    \phi^{n+1/2}_{j+1} 
                                 - 2\phi^{n+1/2}_j 
                                 + \phi^{n+1/2}_{j-1} \right) 
                                   {\delta d_j \over d_j}
       \right].
       \label{eq:accel}
\end{equation}
Generally we find that the potential is so slowly varying that this change is unnecessary.

@@

While there exists an exact solution to this problem, finding it is expensive.  There are four
possible combination of left- and right- traveling shocks and rarefactions,
only one of which is fully consistent with the initial conditions.  Once the
correct physical state is determined, the pressure in the central region can
only be determined by finding the root to an algebraic equation, which is necessarily
an iterative process.  Thus a series of \emph{approximate} Riemann solvers are
typically used.  There are four approximate Riemann solvers in \enzo: two-shock
\citep{toro-1997},
Harten-Lax-van Leer \citep[HLL,][]{toro-1997}, HLL with a contact discontinuity
 \citep[HLLC,][]{toro-1997}, and HLL with
multiple discontinuities \citep[HLLD,][]{Miyoshi05}.  Two-shock is used only
with the PPM method.  HLL and HLLC are used with PPM, MUSCL (both with and
without MHD) and MHD-CT.  HLLD is exclusively an MHD solver, and works with both
MUSCL and MHD-CT methods.

The only approximation that two-shock makes is that both left- and right- moving
waves are shocks.  This solution still requires an iterative method for finding
the pressure in between the two waves.  The HLL method alleviates this iteration
by assuming that there is no central contact discontinuity, and the signal speed
in the central region is approximated by an average over the left- and right-
moving waves.  This method is significantly faster than the two-shock method,
but it is also quite a bit more dissipative.  The HLLC is a three-wave method that improves upon the HLL
method by also including the third wave, the contact discontinuity.
